\Ace: Growing A Statically-Typed Language Inside a Python

### 2014-03-25 ace-oopsla14/string-appendix.tex

\newcommand{\Lagr}{\mathcal{L}}
\newcommand{\strin}[1]{{\tt string\_in #1}}
\newcommand{\lang}[1]{\Lagr\{#1\}}
\newcommand{\stru}[1]{ {\tt string\_union #1}}

@@

\newcommand{\dconvert}{ {\tt dconvert} }
\newcommand{\filter}{ {\tt filter} }
\newcommand{\ifilter}{ {\tt ifilter} }

@@

@@

<t> ::= 			\hfill terms: 					\\
%$x$					\hfill variable 				\alt
%${\tt lambda} x:t$		\hfill abstraction 			\alt
%$t\ t$				\hfill application				\alt
$\filter(\strin[r],t)$ 		\hfill filter substrings			\alt
$[\strin[r]](t)$ \hfill safe conversion \alt
$\dconvert[\strin[r]](t)$ \hfill unsafe conversion

@@

The $\lcs$ language gives static semantics for common regular expression library
functions. In this treatment, we include concatenation and filtering.
The $\filter$ function removes all instances of a regular expression in a string,
while concatenation ($+$) concatenates two strings.

@@

\inferline{T-Convert}
  {\Gamma \vdash [\strin[r]](e):\strin[r]}
  {\Gamma \vdash e : \strin[r'] \\ \lang{r'} \subseteq \lang{r}}

\inferline{T-Dconvert (optional)}
{\Gamma \vdash \dconvert[\strin[r]](e):\strin[r]}
{\Gamma \vdash e : \strin[r']}

\inferline{T-Filter}
  {\Gamma \vdash \filter(\strin[r'], t) : \strin[(r \backslash r') + \emptyset]}
  {\Gamma \vdash t:\strin[r]}

@@

\inferline{E-filterval}
{\filter(\strin[r], e) : \strin[r'] \reduces {\tt rl\_filter}(r,e)}
{ e \val }

\inferline{E-filter}
{\filter(\strin[r], e) : \strin[r'] \reduces  \filter(\strin[r], e' )}     
  { e : T \reduces e' }

\inferline{E-dconvertval (optional) \footnote{See definition ???} }
{\dconvert[\strin[r]](e) : \strin[r'] \reduces {\dconvert[\strin[r]](e'})}
{e:T \reduces e'}

\inferline{E-dconvert (optional) \footnote{See definition ???} }
{\dconvert[\strin[r]](e) : \strin[r'] \ireduces {\tt ifilter}(r, \istrf{e})}
{e \val }

@@

\begin{defn}[Type translation]
  The type translation function $h : Type \rightarrow IType$ is defined as follows:
  \begin{itemize}
    \item $\forall r. h(\strin[r]) = \istr$
    \item $h(\str) = \istr$
  \end{itemize}
\end{defn}

@@

\begin{thm}[Preservation]
  Let $T$ be a type in $\lcs$ and $h(T)=\sigma$ the corresponding type in the internal language.
  For all terms $e$:
  \begin{itemize}
    \item If $e:T$ and $e:T \ireduces i$ then $i : \sigma$ for some $\sigma$.
    \item If $e:T$ and $e:T \reduces e'$ then $e':T$.
  \end{itemize}
\end{thm}
\begin{proof}
The proof is a straightforward induction on the derivation of the combined evaluation relation.
%\begin{itemize}[label=$ $,itemsep=1ex]
%  \item \textbf{E-Ival, E-Ifilterval}. Both cases hold vacuously since the terms at hand are not $\lcs$ terms.
%  \item \textbf{E-String}. Since $e:\str$, it suffices to show that $h(\str) = \istr$ which holds by definition.
%  \item \textbf{E-Sring\_in}. Since $e:\strin[r]$, it suffices to show that $h(\strin[r]) = \istr$ which holds for any $r$ by definition.
%  \item \textbf{E-filter}. Note that ${\tt filter}(\strin[r], e) : \strin[r']$ for some $r'$. 
%    It suffices to show that rl\_filter(r,e)
%    $h(\strin[r]) = \istr$ which holds for any $r$ by definition.
%  \item \textbf{E-Dconvert}. Note that $\dconvert[r](e) : \strin[r]$, so it suffices to show that $h(\strin[r]) = \istr$ which holds by defintiion.
%\end{itemize}
\end{proof}

\begin{thm}[Progress]
  Let $T$ be a type in $\lcs$ and $\sigma$ a type in the internal language.
  For all terms $e$, if $e:T$ then $e:T \reduces^* e' \ireduces^* i$ where $i {\tt ival}$.
\end{thm}
\begin{proof}
By induction on the derivation of $e:T$.
%\begin{itemize}[label=$ $, itemsep=1ex]
%\end{itemize}
\end{proof}

>>>>

\newcommand{\Lagr}{\mathcal{L}}
\newcommand{\strin}{{\tt string\_in}}
\newcommand{\lang}[1]{\Lagr\{#1\}}
\newcommand{\stru}[2]{ {\tt string\_union}(#1,#2)}

@@

\newcommand{\dconvert}[2]{ {\tt dconvert}(#1,#2) }
\newcommand{\filter}[2]{ {\tt filter}(#1,#2) }
\newcommand{\ifilter}[2]{ {\tt ifilter}(#1,#2) }

@@

TODO-nrf Is it ok to make the proofs single-column? If not clean up overflows.

TODO-nrf Make sure failure of unicity isn't causing any problems and write down lemmas about
forms (e.g. $e \in \lang{r} \implies e = \strf{s}$ etc).

@@

<t> ::= 			\hfill terms: 					\\
%$x$					\hfill variable 				\alt
%${\tt lambda} x:t$		\hfill abstraction 			\alt
%$t\ t$				\hfill application				\alt
$\filter{\strin[r]}{t}$ 		\hfill filter substrings			\alt
$[\strin[r]](t)$ \hfill safe conversion \alt
$\dconvert{\strin[r]}{t}$ \hfill unsafe conversion

@@

The $\lcs$ language gives static semantics for common regular expression library
functions. In this treatment, we include concatenation and filtering.
The ${\tt filter}$ function removes all instances of a regular expression in a string,
while concatenation ($+$) concatenates two strings.

@@

\inferline{T-Convert}
{\Gamma \vdash [\strin[r']](e):\strin[r']}
{\Gamma \vdash e : \strin[r] \\ \lang{r'} \subseteq \lang{r}}

\inferline{T-Dconvert (optional)}
{\Gamma \vdash \dconvert{\strin[r']}{e}:\strin[r']}
{\Gamma \vdash e : \strin[r]}

\inferline{T-Filter}
{\Gamma \vdash \filter{\strin[r']}{t} : \strin[(r \backslash r') + \emptyset]}
{\Gamma \vdash t:\strin[r]}

@@

\inferline{E-filterval}
{\filter{\strin[r']}{e} : \strin[r\backslash r' + \emptyset] \reduces {\tt rl\_filter}(r,e)}
{ e \val }

\inferline{E-filter}
{\filter{\strin[r']}{e} : \strin[r\backslash r' + \emptyset] \reduces  \filter{\strin[r]}{ e' }}     
{ e : \strin[r] \reduces e' }

\inferline{E-convertval}
{ [\strin[r']](e) : \strin[r'] \reduces [\strin[r']](e') }
{ e : \strin[r] \reduces e' }

\inferline{E-convert}
{ [\strin[r']](e) : \strin[r'] \ireduces \istrf{e} }
{ e \val }

\inferline{E-dconvertval (optional) \footnote{See definition ???} }
{\dconvert{\strin[r']}{e} : \strin[r'] \reduces {\dconvert{\strin[r']}{e'}}}
{e:\strin[r] \reduces e'}

\inferline{E-dconvert (optional) \footnote{See definition ???} }
{\dconvert{\strin[r']}{e} : \strin[r'] \ireduces \ifilter{r'}{\istrf{e}}}
{e \val }

@@

\begin{defn}[Type Translation Function $h$]
  The type translation function $h : Type \rightarrow IType$ is defined as follows:
  \begin{itemize}
    \item $\forall r. h(\strin[r]) = \istr$
    \item $h(\str) = \istr$
  \end{itemize}
\end{defn}

@@

Before presenting progress and preservation theorems, we establish some definitions
and lemmas necessary for the proofs.

\begin{defn}[Types of internal values]
The types of internal values are defined as follows:
\begin{itemize}
  \item If $e=\istrf{s}$ then $e:\istr$.
  \item If $e=\ifilter(r, \istrf{s})$ then $e:\istr$.
  \item If $e=\istrf{s_1}+\istrf{s_2}$ then $e:\istr$.
\end{itemize}
\end{defn}

Definition 2 is an additional assumption about
the internal language. To summarize, we assume an internal language containing 
strings together with operations for string concatentation and filtering. We
expect closure over strings for both operations.
Recall that $\ifilter$ is only needed for dynamic casts, which may be removed without
descreasing the expressivity or even usability of the language.

%\begin{lem}[Canonical Forms for $\lcs$]
%We identify only canonical forms necessary for the type safety proof.
%\begin{itemize}
%  \item If $e \val$ and $e:\str$ then $e = \strf{s}$.
%\end{itemize}
%\end{lem}

\begin{thm}[Preservation]
  Let $T$ be a type in $\lcs$ and $h(T)=\sigma$ the corresponding type in the internal language.
  For all terms $e$:
  \begin{itemize}
    \item If $e:T$ and $e:T \ireduces i$ then $i : \sigma$ for some $\sigma$.
    \item If $e:T$ and $e:T \reduces e'$ then $e':T$.
  \end{itemize}
\end{thm}
\begin{proof}
The proof is a straightforward induction on the derivation of the combined evaluation relation.
\begin{itemize}[label=$ $,itemsep=1ex]
  \item \textbf{E-Ival, E-Ifilterval}. Both cases hold since the terms at hand are not $\lcs$ terms.
  \item \textbf{E-Stringval, E-strval}. Both cases hold since no reduction is possible. 
  \item \textbf{E-String}. By the definition of typing for internal terms, $\istrf{e} : \istr$. It suffices to show that $h(\str) = \istr$, which follows from the definition of $h$.
  \item \textbf{E-String\_in}. By the definition of typing for internal terms, $\istrf{e} : \istr$. It suffices to show that $h(\strin[r]) = \istr$, which follows from the definition of $h$ for arbitrary $r$.
  \item \textbf{E-concat}. By the definition of typing for internal terms, $\istrf{e_1} + \istrf{e_2} : \istr$. So it suffices to show that $h(\strin[r_1+r_2]) = \istr$, which follows from the definition of $h$ for arbitrary $r_1$, $r_2$.
  \item \textbf{E-Filterval}. We have that $\filter{\strin[r']}{e}:\strin[r\backslash r' + \emptyset]$. By inversion of T-Filter, $e:\strin[r]$.
    By T-Equiv-string\_in (which is bidirectional), $e \in \lang{r}$.
    By theorem 2, ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$.
    By T-Equiv-string\_in,  $e \in \lang{r}$ and ${\tt rl\_filter}(r,e) \in \lang{r\backslash r' + \emptyset}$ implies ${\tt rl\_filter}(r,e):\strin[r\backslash r' + \emptyset]$.
  \item \textbf{E-Filter}. By inversion, $e : \strin[r'] \reduces e'$ so by the induction $e':\strin[r']$.
    Therefore, ${\tt filter}(\strin[r], e') : \strin[r \backslash r' + \emptyset]$ by T-Filter.
  \item \textbf{E-Convertval}. It suffices to show that $h(\strin[r]) = \istr$, which is true by definition.
  \item \textbf{E-Convert}. By inversion and induction, $e' : \strin[r]$. We know that $[\strin[r']](e) : \strin[r']$, so by inversion of T-Convert $\lang{r'} \subseteq \lang{r}$.
    It follows that $[\strin[r']](e') : \strin[r']$.
  \item \textbf{E-DConvert}. By inversion, $e : \strin[r'] \reduces e'$. By the induction hypothesis, $e' : \strin[r']$; therefore, by T-Dconvert, $\dconvert[\strin[r]](e):\strin[r]$.
  \item \textbf{E-DConvertval}. By the definition of typing for internal terms, $\ifilter(r,\istrf{e}) : \istr$. It suffices to show that $h(\strin[r_1+r_2]) = \istr$, which follows from the definition of $h$.
\end{itemize}
\end{proof}

\begin{thm}[Progress]
  If $e:T$ then $e:T \reduces^* e' \ireduces^* i$ where $i \ival$.
\end{thm}
\begin{proof}
By induction on the derivation of $e:T$.
\begin{itemize}[label=$ $,itemsep=1ex]
  \item \textbf{T-Equiv-Include}. Since $e \in \lang{r}$, $e = \strf{s}$ for some $s$; therefore, $e \val$ by E-Strinval.
  \item \textbf{T-Minimum}. 
\end{itemize}
\end{proof}
